Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-12 paper/paper.tex
update weights

\begin{abstract}
We introduce a method that uses generic programming tensor network to compute various properties of a class of hard graph problems that can be quantified locally.
Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph.
Other hard graph problems solvable in our framework includes the cutting problem, the K-coloring problem, the maximal clique problem, the dominating set problem, and satisfiability problem, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
We show how our method helps people understand these hard problems better.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there is an important class of problems that can be quantified locally,
including the independent set problem, the cutting problem, the K-coloring problem, the satisfiability problem,
the maximal clique problem, and the vertex cover problem et al ~\cite{Moore2011}.
There is a widespread interest of these problem because they have wide range of applications in scheduling, logistics,
wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
\blue{need more diversed applications not limited to IS}

The most studied version of the above problems is finding the maximum/minimum set size and one of the sets of that size satisfying the contraints,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. it is unlikely to solve them in a polynomial time.
In this paper, we focus on the less studied \textit{property} computation which can be harder but is crutial towards understanding a hard problem.
Here, the \textit{property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version of the above problems, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.

We propose to view the computation of all these properties from an abstract algebra perspective and use a unified framework,
generic tensor-network, to solve these seeming unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network or einsum in different fields,
where its tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
Given a good contraction order, the time complexity of contracting a tensor network is $2^{{\rm tw}(G)}$~\cite{Markov2008}, where ${\rm tw}(G)$ is the treewidth of graph $G$.
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time.
Although this time complexity is just on par with the dynamic programming algorithms~\cite{Courcelle1990, Fomin2013} for solving the maximum size and one of the best set, tensor network has richer algebraic structure, and this structure is crutial for property computation.
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

%The independent set problem is related to physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
%in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
%Property computation can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007},
%to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
%However, there is a lack of general and versatile tool to compute these different properties of independent sets.

To avoid the discussion being unnecessaryly diversed, we mainly focus on the independent set problem,
and discuss other combinatorial optimization problems in \App{sec:otherproblems}.
The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce an independent set problem to a tensor network contraction problem \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting},\Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.

@@

\section{Generic programming tensor contractions}\label{sec:generic}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017},
the elements in the tensor networks are limited to standard number types such as floating point numbers and integers.
Owning to the development of modern compiling technology, we no longer need to re our imagination to standard number types.
One of the key concepts that push the technology forward is called generic programming:

@@

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Property to compute} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their counting} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Sum Product Tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with Bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with Set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section is about reducing the independent set problem to a tensor network contraction problem,
for people with physics background, it is suggested to use an alternative approach described in \App{sec:energymodel} to understand this construction from the partition function perspective.
Let $G=(V,E)$ be a weighted undirected graph with each vertex $i\in V$ being associated with a weight $w_i$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$,
and we denote the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$. 
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i, w_i)$ as
\begin{equation}
    W(x_i, w_i) = \left(\begin{matrix}
        1 \\
        x_i^{w_i}
    \end{matrix}\right).\label{eq:vertextensor}
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i, w_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i, w_i)_1=x_i^{w_i}$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1^{w_1}, x_{2}^{w_2}, \ldots,x_{|V|}^{w_{|V|}}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i, w_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.

@@

In the graphical representation of tensor network on the right panel,
we use a circle to represent a tensor, a hyperedge in cyan color to represent the degree of freedom $s_a$,
and a hyperedge in red color to represent a degree of freedom $s_b$.
Tensors sharing a same degree of freedom are connected by that hyperedge.
The contraction of this tensor network can be evaluated manually:
\begin{equation}
    \begin{split}
    P({\rm K2},\{x_a^{w_a}, x_b^{w_b}\}) &=
    \left(\begin{matrix} 1 & x_a^{w_a} \end{matrix}\right)
    \left(\begin{matrix} 1 & 1\\ 1  & 0 \end{matrix}\right)
    \left(\begin{matrix} 1 \\ x_b^{w_b} \end{matrix}\right)\\
    &= 1 + x_a^{w_a} + x_b^{w_b}
    \end{split}
\end{equation}
The resulting polynomial represents 3 different independent sets $\{\}$, $\{a\}$ and $\{b\}$, with weights being $0$, $w_a$ and $w_b$ respectively.
\end{example}

@@

\section{Independence polynomial}\label{sec:indpoly}
In the special case of $x_i = x$ and $w_i = 1$, the tensor network contraction result directly corresponds to the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}.
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in graph $G=(V,E)$. The total number of independent sets is thus equal to $I(G, 1)$.
Its connection to \Eq{eq:idp} can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise. The summation counts the number of independent sets of size $k$. 
One can feed the standard computer numbers types into this tensor network and evaluate the independence polynomial by contracting it numerically.
However, we are more interested in finding the coefficients of this polynomial, because this quantity tells us the counting of independent sets at each size.
To achieve this, let us first elevate the tensor elements $0$s and $1$s in tensors $W(x, 1)$ and $B$ from integers and floating point numbers to the additive identity,
$\mymathbb{0}$, and multiplicative identity, $\mymathbb{1}$, of a commutative semiring as discussed in Sec.~\ref{sec:generic}.
Let us create a polynomial type and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a coefficient vector $a = (a_0, a_1, \ldots, a_k) \in \mathbb{R}^k$, so, e.g., $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{equation}
    \eqname{PN}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, a_{2}b_{0} + a_{1}b_{1} + a_{0}b_{2},  \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),  \\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{equation}
Here, the multiplication operation can be evaluated efficiently using the convolution theorem~\cite{Schonhage1971}.
These operations are standard addition and multiplication operations of polynomials, and the polynomial type forms a commutative ring. The tensors $W$ and $B$ can thus be written as 
\begin{equation}
    W^{\rm PN} = \left(\begin{matrix}
        \mymathbb{1} \\
        (0,1)
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm PN} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
By contracting the tensor network with the polynomial type, we have the exact representation of the independence polynomial.
In practise, using the polynomial type suffers a space overhead proportional to $\alpha(G)$ because each polynomial requires a vector of such size to store the coefficients. 
In \App{sec:finitefield}, we provide an accurate and memory efficient method to find the polynomial by fitting it in a finite field.
Alternatively, one can use the method in \App{sec:fft} to fit this polynomial more time efficiently with complex number at the cost of precision.

@@

\subsection{Sampling extremely large configuration space}
When the graph size goes large, the configurations might be impossible to fit into any type of storage.
Then instead of using the set algebra, one can use a binary sum-product tree as a compact representation of the configurations.
We use a quadruple $(type, data, left, right)$ to represent a node in the tree, where $type$ is one of
$\{\texttt{LEAF}, \texttt{ZERO}, \texttt{SUM}, \texttt{PROD}\}$, $data$ is a bit string as the content in a \texttt{LEAF} node,
$left$ and $right$ are left and right operands of \texttt{SUM} and \texttt{PROD} nodes.

>>>>

\begin{abstract}
We introduce a method that uses generic programming tensor network to compute various properties of independent sets,
which include, for example, the size of the maximum independent sets and the number and enumeration of independent sets of a given size.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph.
Our framework can be easily extended to compute properties of other problems such as the cut size, coloring, and maximal cliques, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there is an important class of problems that can be quantified locally,
including the independent set problem, the cutting problem, the 3-coloring problem, the 3-satisfiability problem,
the maximal clique problem, and the vertex cover problem et al ~\cite{Moore2011}.
There is a widespread interest of these problem because they have wide range of applications in scheduling, logistics,
wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
\blue{need more diversed applications not limited to IS}

The most studied version of the above problems is finding the maximum/minimum set size and one of the solution with that size satisfying the contraints,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. there is unlikely a polynomial time algorithm for solving them.
In this paper, we focus on the less studied \textit{property} computation of these problems which can be harder but is crutial towards understanding a problem.
Here, the \textit{property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.

We propose to view the computation of all these properties from an abstract algebra perspective and use a unified framework,
generic tensor-network, to solve these seeming unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network or einsum in different fields,
where its tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
%Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
Given a good contraction order, contracting a tensor network in general has time complexity $2^{{\rm tw}(G)}$~\cite{Markov2008}, where ${\rm tw}(G)$ is the treewidth of graph $G$.
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time.
Although this time complexity is just on par with the dynamic programming algorithms~\cite{Courcelle1990, Fomin2013} for solving the maximum size and one of the best set, tensor network has richer algebraic structure, and this structure is crutial for computing properties.
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

To avoid the discussion being overly diversed, we limit the target problem to the independent set problem,
and discuss other combinatorial optimization problems in \App{app:otherproblems}.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$,
and we denote the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$. 
%The independent set problem is related to physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
%in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
%Property computation can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007},
%to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}.
%However, there is a lack of general and versatile tool to compute these different properties of independent sets.

The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce an independent set problem to a tensor network contraction problem \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting},\Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{app:tensorbad}.

@@

\section{Generic programming tensor contractions}\label{sec:generic}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017},
the elements in the tensor networks are limited to standard number types such as floating point numbers and integers.
Owning to the development of modern compiling technology, we no longer need to limit our imagination to standard number types.
One of the key concepts that push the technology forward is called generic programming:

@@

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Property to compute} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended Tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their counting} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Sum Product Tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with Bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with Set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section is about reducing the independent set problem to a tensor network contraction problem,
for people with physics background, it is suggested to use an alternative approach described in \App{sec:energymodel} to understand this construction and the independence polynomial from the partition function perspective.
Let $G=(V,E)$ be a weighted graph with each vertex $i\in V$ being associated with a weight $w_i$.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i, w_i)$ as
\begin{equation}
    W(x_i, w_i) = \left(\begin{matrix}
        1 \\
        x_i^{w_i}
    \end{matrix}\right).\label{eq:vertextensor}
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i, w_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i, w_i)_1=x_i^{w_i}$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1^{w_1}, x_{2}^{w_2}, \ldots,x_{|V|}^{w_{|V|}}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i, w_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.

@@

In the graphical representation of tensor network on the right panel,
we use a circle to represent a tensor, a hyperedge in cyan color to represent the degree of freedom $s_a$,
and a hyperedge in red color to represent a degree of freedom $s_b$.
Tensors sharing a same degree of freedom are connected by that hyperedge.
The contraction of this tensor network can be evaluated manually:
\begin{equation}
    \begin{split}
    P(K2,\{x_a^{w_a}, x_b^{w_b}\}) &=
    \left(\begin{matrix} 1 & x_a^{w_a} \end{matrix}\right)
    \left(\begin{matrix} 1 & 1\\ 1  & 0 \end{matrix}\right)
    \left(\begin{matrix} 1 \\ x_b^{w_b} \end{matrix}\right)\\
    &= 1 + x_a^{w_a} + x_b^{w_b}
    \end{split}
\end{equation}
The resulting polynomial represents 3 different independent sets $\{\}$, $\{a\}$ and $\{b\}$, with weights being $0$, $w_a$ and $w_b$ respectively.
\end{example}

@@

\section{Independence polynomial}\label{sec:indpoly}
In the special case of $x_i = x$ and $w_i = 1$, the tensor network contraction result directly corresponds to the independence polynomial.
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in graph $G=(V,E)$. The total number of independent sets is thus equal to $I(G, 1)$.
Its connection to \Eq{eq:idp} can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise. The summation counts the number of independent sets of size $k$. 
One can feed the standard computer numbers types into this tensor network and evaluate the independence polynomial by contracting it numerically.
However, we are more interested in finding the coefficients of this polynomial, because this quantity tells us the counting of independent sets at each size.
To achieve this, let us first elevate the tensor elements $0$s and $1$s in tensors $W(x, 1)$ and $B$ from integers and floating point numbers to the additive identity,
$\mymathbb{0}$, and multiplicative identity, $\mymathbb{1}$, of a commutative semiring as discussed in Sec.~\ref{sec:generic}.
Let us create a polynomial type and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a coefficient vector $a = (a_0, a_1, \ldots, a_k) \in \mathbb{R}^k$, so, e.g., $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{equation}
    \eqname{PN}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, a_{2}b_{0} + a_{1}b_{1} + a_{0}b_{2},  \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),  \\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{equation}
Here, the multiplication operation can be evaluated efficiently using the convolution theorem~\cite{Schonhage1971}.
These operations are standard addition and multiplication operations of polynomials, and the polynomial type forms a commutative ring. The tensors $W$ and $B$ can thus be written as 
\begin{equation}
    W^{\rm PN} = \left(\begin{matrix}
        \mymathbb{1} \\
        (0,1)
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm PN} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
By contracting the tensor network with the polynomial type, we have the exact representation of the independence polynomial.
In practise, using the polynomial type suffers a space overhead proportional to $\alpha(G)$ because each polynomial requires a vector of such size to store the coefficients. 
Here, we propose to find the independence polynomial by fitting $\alpha(G)+1$ random pairs of $x_{i}$ and $y_{i} = I(G,x_{i})$. One can then compute the independence polynomial coefficients $a_{i}$ by solving the linear equation: 
\begin{equation}
\left(\begin{matrix}
1 & x_0 & x_0^2 & \ldots & x_0^{\alpha(G)} \\
1 & x_1 & x_1^2 & \ldots & x_1^{\alpha(G)} \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_{\alpha(G)} & x_{\alpha(G)}^2 & \ldots & x_{\alpha(G)}^{\alpha(G)}
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_{\alpha(G)}
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_{\alpha(G)}
\end{matrix}\right).\label{eq:lineareq}
\end{equation}
With this approach, we do not incur the linear overhead in space. However, because the independence polynomial coefficients can have a huge order-of-magnitude range, if we use floating point numbers in the computation, the round-off errors can be significant for the counting of large-size independent sets.
In addition, the number could easily overflow if we use fixed-width integer types.
The big integer type is also not a good option because big integers with varying width can be very slow and is incompatible with GPU devices. These problems can be solved by introducing a finite-field algebra $\text{GF}(p)$:
\begin{equation}
\eqname{GF$(p)$}
\begin{split}
    x ~\oplus~ y &= x+y\pmod p,\\
    x ~\odot~ y &= xy\pmod p,\\
    \mymathbb{0} &= 0,\\
    \mymathbb{1} &= 1.
\end{split}\label{eq:finitefield}
\end{equation}
With a finite-field algebra, we have the following observations:
\begin{enumerate}
    \item One can use Gaussian elimination~\cite{Golub2013} to solve the linear equation \Eq{eq:lineareq} since it is a generic algorithm that works for any elements with field algebra. The multiplicative inverse of a finite-field algebra can be computed with the extended Euclidean algorithm.
    \item Given the remainders of a larger unknown integer $x$ over a set of co-prime integers $\{p_1, p_2, \ldots, p_n\}$,
    $x \pmod {p_1 \times p_2 \times \ldots \times p_n}$ can be computed using the Chinese remainder theorem. With this, one can infer big integers from small integers.
\end{enumerate}
With these observations, we develop Algorithm~\ref{alg:finitefield} to compute the independence polynomial exactly without introducing space overheads.
The algorithm iterates over a sequence of large prime numbers until convergence.
In each iteration, we choose a large prime number $p$, and contract the tensor networks to evaluate the polynomial for each variable $\chi = (x_{0}, x_{1}, \ldots, x_{\alpha(G)})$ on ${\rm GF}(p)$ and denote the outputs as $(y_0, y_1, \ldots, y_{\alpha(G)}) \pmod p$.
Then we solve \Eq{eq:lineareq} using Gaussian elimination on ${\rm GF}(p)$ to find the coefficient modulo $p$, $A_p \equiv (a_0, a_1, \ldots, a_{\alpha(G)})\pmod p$.
As the last step of each iteration, we apply the Chinese remainder theorem to update $A \pmod P $ to $ A \pmod {P\times p}$, where $P$ is a product of all prime numbers chosen in previous iterations.
If this number does not change compared with the previous iteration, it indicates the convergence of the result and the program terminates.
All computations are done with integers of fixed width $W$ except the last step of applying the Chinese remainder theorem, where we use arbitrary precision integers to represent the counting.
In \App{app:fft}, we provide another method to solve the linear equation using discrete Fourier transformation.

@@

\subsection{Sampling extremely large configuration space}
In many times, the number of configurations is too large to fit into any type of storage.
Then instead of using the set algebra, we can use a binary sum-product tree as a compact representation of the results.
We use a quadruple $(type, data, left, right)$ to represent a node in the tree, where $type$ is one of
$\{\texttt{LEAF}, \texttt{ZERO}, \texttt{SUM}, \texttt{PROD}\}$, $data$ is a bit string as the content is a \texttt{LEAF} node,
$left$ and $right$ are left and right operands of \texttt{SUM} and \texttt{PROD} nodes.
